---
title: "PCA Analysis — Rocket League 3v3"
subtitle: "Dimensionality Reduction of ballchasing.com Match Data"
author: "Curtis Mesmer, David Meyers, Mia Lewis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)
```

# Libraries

```{r libraries}
library(Hmisc)
library(tidyverse)
library(skimr)
library(DataExplorer)
library(DescTools)
library(tidyverse)
library(factoextra)
```

# Load Data

Two files exported from `RL_Data_Preparation.Rmd`:

- `pca_data.csv` — 58 numeric features only (input to PCA)
- `working.csv` — full working dataset with rank, match result, etc.

```{r load-data}
pca_data <- read_csv("pca_data.csv", show_col_types = FALSE)
working  <- read_csv("working.csv", show_col_types = FALSE)

rank_labels <- c(
  "bronze-1", "bronze-2", "bronze-3",
  "silver-1", "silver-2", "silver-3",
  "gold-1", "gold-2", "gold-3",
  "platinum-1", "platinum-2", "platinum-3",
  "diamond-1", "diamond-2", "diamond-3",
  "champion-1", "champion-2", "champion-3",
  "grand-champion-1", "grand-champion-2", "grand-champion-3"
)

working <- working %>%
  mutate(rank_factor = factor(rank_tier,
                              levels = 1:21,
                              labels = rank_labels,
                              ordered = TRUE),
         match_result = as.factor(match_result))
```

## Verification

```{r verify}
cat("PCA data:", nrow(pca_data), "rows x", ncol(pca_data), "columns\n")
cat("Working data:", nrow(working), "rows x", ncol(working), "columns\n")
cat("Row counts match:", nrow(pca_data) == nrow(working), "\n")
cat("Any NAs in PCA data:", sum(is.na(pca_data)), "\n")
cat("All numeric:", all(sapply(pca_data, is.numeric)), "\n")
```

# Data Exploration
```{r}
skim_without_charts(working)
```

```{r}
cat("Match result distribution:\n")
table(working$match_result)

cat("\nTeam colour distribution:\n")
table(working$team_color)

cat("\nRank tier distribution:\n")
table(working$rank_tier, useNA = "ifany")
```

```{r}
# Total obs
nrow(working)  # 76,035

# Match results
working %>%
  group_by(match_result) %>%
  summarise(n = n(), pct = round(100 * n / sum(n), 1))

# Rank distribution
working %>%
  group_by(rank_name) %>%
  summarise(n = n(), pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))

# Rank groups (if you have rank_group variable)
working %>%
  group_by(rank_group) %>%
  summarise(n = n(), pct = round(100 * n / sum(n), 1))
```

# Principal Component Analysis

## Run PCA

```{r run-pca}
pca_result <- prcomp(pca_data, center = TRUE, scale. = TRUE)
```

## Variance Explained

```{r variance-table}
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
cum_var       <- cumsum(var_explained)

var_table <- tibble(
  PC             = paste0("PC", 1:ncol(pca_data)),
  Eigenvalue     = round(pca_result$sdev^2, 4),
  Pct_Variance   = round(var_explained * 100, 2),
  Cumulative_Pct = round(cum_var * 100, 2)
)

print(var_table, n = 25)
```

```{r variance-thresholds}
cat("PCs needed for 70% variance:", which(cum_var >= 0.70)[1], "\n")
cat("PCs needed for 80% variance:", which(cum_var >= 0.80)[1], "\n")
cat("PCs needed for 90% variance:", which(cum_var >= 0.90)[1], "\n")
```

## Scree Plot

```{r scree-plot, fig.cap="Figure 1: Scree plot — variance explained by each principal component."}
fviz_eig(pca_result,
         ncp = 20,
         addlabels = TRUE,
         ylim = c(0, 25),
         main = "Scree Plot — Variance Explained per PC")
```

## Correlation with Rank

Pearson correlation between each PC score vector and the ordinal rank variable (1–21).

```{r rank-correlations}
rank_numeric <- as.numeric(working$rank_factor)

rank_cors <- tibble(
  PC = paste0("PC", 1:10),
  Correlation_with_Rank = round(sapply(1:10, function(i) {
    cor(pca_result$x[, i], rank_numeric, use = "complete.obs")
  }), 4),
  Abs_Correlation = round(abs(Correlation_with_Rank), 4),
  Interpretation = case_when(
    Abs_Correlation > 0.5  ~ "Strong rank effect",
    Abs_Correlation > 0.3  ~ "Moderate rank effect",
    Abs_Correlation > 0.15 ~ "Weak rank effect",
    TRUE                   ~ "Minimal rank effect"
  )
)

print(rank_cors)
```
## Correlation with Rank

Pearson correlation between each PC score vector and the ordinal win variable (1-2).

```{r win-correlations}
win_numeric <- as.numeric(working$match_result)

win_cors <- tibble(
  PC = paste0("PC", 1:10),
  Correlation_with_Win = round(sapply(1:10, function(i) {
    cor(pca_result$x[, i], win_numeric, use = "complete.obs")
  }), 4),
  Abs_Correlation = round(abs(Correlation_with_Win), 4),
  Interpretation = case_when(
    Abs_Correlation > 0.5  ~ "Strong win effect",
    Abs_Correlation > 0.3  ~ "Moderate win effect",
    Abs_Correlation > 0.15 ~ "Weak win effect",
    TRUE                   ~ "Minimal win effect"
  )
)

print(win_cors)
```

## Top Loadings per PC

```{r top-loadings}
for (i in 1:10) {
  cat("\n--- PC", i,
      "(", round(var_explained[i] * 100, 1), "% variance",
      "| rank r =", round(rank_cors$Correlation_with_Rank[i], 3),
      ") ---\n")

  loadings_i <- pca_result$rotation[, i]
  sorted <- sort(loadings_i, decreasing = TRUE)

  cat("\n  Top 5 POSITIVE loadings:\n")
  top_pos <- head(sorted, 5)
  for (j in seq_along(top_pos)) {
    cat("    ", formatC(names(top_pos)[j], width = -55),
        formatC(round(top_pos[j], 4), width = 8, format = "f"), "\n")
  }

  cat("\n  Top 5 NEGATIVE loadings:\n")
  top_neg <- tail(sorted, 5)
  for (j in seq_along(top_neg)) {
    cat("    ", formatC(names(top_neg)[j], width = -55),
        formatC(round(top_neg[j], 4), width = 8, format = "f"), "\n")
  }
}
```

## Variable Contribution Plots

```{r contrib-plots, fig.height=5}
for (i in 1:6) {
  p <- fviz_contrib(pca_result, choice = "var", axes = i, top = 15,
                    title = paste0("Top 15 Variables Contributing to PC", i))
  print(p)
}
```

## Component Descriptions

Labels derived directly from loading structure. PCs 1–6 are described based on their top positive and negative loadings. PCs 7–16 are retained for the 80% variance threshold but not individually interpreted, as their lower variance contributions and increasingly complex loading structures make substantive labelling unreliable (Figueiredo et al., 2020).

```{r component-descriptions}
labels_table <- tibble(
  PC = paste0("PC", 1:10),
  Pct_Variance = round(var_explained[1:10] * 100, 1),
  Rank_Cor = round(rank_cors$Correlation_with_Rank, 3),
  Description = c(
    "Movement Speed & Boost Collection (+ slow speed, defensive position, ground | - boost collected, boost stolen, big pads)",
    "Boost Reserves & Ball Distance (+ avg boost amount, overfill, distance to ball | - low boost %, closest to ball)",
    "Defensive vs. Offensive Field Position (+ defensive third/half, shots against | - offensive third/half)",
    "Ball & Teammate Proximity (+ closest to ball, slow speed, score | - distance to ball, distance to mates)",
    "Big Pad vs. Small Pad Boost Collection (+ full boost %, high boost bands | - small pad counts, small pad amounts)",
    "Scoring Output (+ behind ball, goals, score, shooting % | - in front of ball, goals against)",
    "PC7 — retained for variance, not individually interpreted",
    "PC8 — retained for variance, not individually interpreted",
    "PC9 — retained for variance, not individually interpreted",
    "PC10 — retained for variance, not individually interpreted"
  )
)

print(labels_table, width = Inf)
```

## Biplot

```{r biplot, fig.width=12, fig.height=10, fig.cap="Figure 2: PCA biplot — variable loadings overlaid on PC1 vs PC2 scores."}
fviz_pca_biplot(pca_result,
                label = "var",
                alpha.ind = 0.03,
                col.var = "firebrick",
                repel = TRUE,
                title = "PCA Biplot — PC1 vs PC2 with Variable Loadings")
```


# PCA Score Visualisations

```{r pca-scores-setup}
assign_rank_group <- function(rf) {
  case_when(
    rf %in% c("bronze-1","bronze-2","bronze-3",
              "silver-1","silver-2","silver-3",
              "gold-1","gold-2","gold-3",
              "platinum-1","platinum-2","platinum-3")                   ~ "Below Diamond",
    rf %in% c("diamond-1","diamond-2","diamond-3")                      ~ "Diamond",
    rf %in% c("champion-1","champion-2","champion-3")                   ~ "Champion",
    rf %in% c("grand-champion-1","grand-champion-2","grand-champion-3") ~ "Grand Champion",
    TRUE ~ "Other"
  )
}

pca_scores <- as.data.frame(pca_result$x[, 1:10]) %>%
  mutate(
    rank_factor  = working$rank_factor,
    match_result = working$match_result,
    rank_group   = factor(assign_rank_group(rank_factor),
      levels = c("Below Diamond", "Diamond", "Champion", "Grand Champion"),
      ordered = TRUE)
  )
```

## PC1 vs PC2 — Coloured by Rank

```{r pc1-pc2-rank, fig.cap="Figure 3: PC1 vs PC2 coloured by rank."}
ggplot(pca_scores, aes(x = PC1, y = PC2, color = rank_factor)) +
  geom_point(alpha = 0.05, size = 0.3) +
  labs(title = "PC1 vs PC2 — Coloured by Rank",
       x = paste0("PC1: Movement Speed & Boost Collection (", round(var_explained[1] * 100, 1), "%)"),
       y = paste0("PC2: Boost Reserves & Ball Distance (", round(var_explained[2] * 100, 1), "%)"),
       color = "Rank") +
  theme_minimal() +
  theme(legend.position = "right")
```

## PC1 vs PC2 — Coloured by Match Result

```{r pc1-pc2-result, fig.cap="Figure 4: PC1 vs PC2 coloured by match result."}
ggplot(pca_scores, aes(x = PC1, y = PC2, color = match_result)) +
  geom_point(alpha = 0.05, size = 0.3) +
  labs(title = "PC1 vs PC2 — Coloured by Match Result",
       x = paste0("PC1: Movement Speed & Boost Collection (", round(var_explained[1] * 100, 1), "%)"),
       y = paste0("PC2: Boost Reserves & Ball Distance (", round(var_explained[2] * 100, 1), "%)"),
       color = "Result") +
  theme_minimal()
```

## PC1 Distribution by Rank Group

```{r pc1-boxplot, fig.cap="Figure 5: PC1 score distribution by rank group."}
ggplot(pca_scores, aes(x = rank_group, y = PC1, fill = rank_group)) +
  geom_boxplot(outlier.alpha = 0.1, outlier.size = 0.3) +
  labs(title = "PC1 Score Distribution by Rank Group",
       x = "Rank Group", y = "PC1 (Movement Speed & Boost Collection)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

## PC3 Distribution by Rank Group

```{r pc3-boxplot, fig.cap="Figure 6: PC3 score distribution by rank group."}
ggplot(pca_scores, aes(x = rank_group, y = PC3, fill = rank_group)) +
  geom_boxplot(outlier.alpha = 0.1, outlier.size = 0.3) +
  labs(title = "PC3 Score Distribution by Rank Group",
       x = "Rank Group", y = "PC3 (Defensive vs. Offensive Field Position)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

## PC1 vs PC3 — Coloured by Result

```{r pc1-pc3-result, fig.cap="Figure 7: PC1 vs PC3 coloured by match result."}
ggplot(pca_scores, aes(x = PC1, y = PC3, color = match_result)) +
  geom_point(alpha = 0.05, size = 0.3) +
  labs(title = "PC1 vs PC3 — Coloured by Match Result",
       x = paste0("PC1: Movement Speed & Boost Collection (", round(var_explained[1] * 100, 1), "%)"),
       y = paste0("PC3: Defensive vs. Offensive Field Position (", round(var_explained[3] * 100, 1), "%)"),
       color = "Result") +
  theme_minimal()
```

## Rank Group Counts

```{r rank-counts}
pca_scores %>%
  count(rank_group) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  print()
```

## PC Score Means by Rank Group

```{r pc-means-by-rank}
pca_scores %>%
  group_by(rank_group) %>%
  summarise(
    n = n(),
    across(PC1:PC6, ~ round(mean(.), 2)),
    .groups = "drop"
  ) %>%
  print(width = Inf)
```

## PC Score Means by Match Result

```{r pc-means-by-result}
pca_scores %>%
  group_by(match_result) %>%
  summarise(
    n = n(),
    across(PC1:PC6, ~ round(mean(.), 2)),
    .groups = "drop"
  ) %>%
  print(width = Inf)
```

## PC Score Means by Rank Group and Match Result

```{r pc-means-rank-result}
pca_scores %>%
  group_by(rank_group, match_result) %>%
  summarise(
    n = n(),
    across(PC1:PC6, ~ round(mean(.), 3)),
    .groups = "drop"
  ) %>%
  print(width = Inf, n = Inf)
```


# Summary

```{r summary}
cat("=================================================================\n")
cat("PCA SUMMARY\n")
cat("=================================================================\n\n")

cat("Input: ", ncol(pca_data), "standardised features\n")
cat("Observations: ", nrow(pca_data), "\n")
cat("Method: prcomp(center = TRUE, scale. = TRUE) — unrotated\n\n")

cat("Variance thresholds:\n")
cat("  70%:", which(cum_var >= 0.70)[1], "PCs\n")
cat("  80%:", which(cum_var >= 0.80)[1], "PCs\n")
cat("  90%:", which(cum_var >= 0.90)[1], "PCs\n\n")

cat("PC1 rank correlation: r =", round(rank_cors$Correlation_with_Rank[1], 3), "(moderate)\n")
cat("PCs 2-6 rank correlations: |r| <", round(max(rank_cors$Abs_Correlation[2:6]), 3), "\n")
cat("  → PCs 2-6 capture behavioural variation largely independent of rank\n\n")

cat("Top 6 PCs account for",
    round(cum_var[6] * 100, 1), "% of total variance\n")
cat("16 PCs account for",
    round(cum_var[16] * 100, 1), "% of total variance\n")
```


# Session Info

```{r session-info}
sessionInfo()
```

```{r run-all}

# From your RMD output or re-running:
names(working)
```
